---
title: "Simple animation with anim.plots"
author: "David Hugh-Jones"
date: "13/05/2017"
output:
  ioslides_presentation:
    fig_height: 2.5
    fig_width: 4
    self_contained: no
    smaller: yes
---

<link rel="stylesheet" href="http://vis.supstat.com/assets/themes/dinky/css/scianimator.css">
<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.7.1/jquery.min.js"></script>
<script src="http://vis.supstat.com/assets/themes/dinky/js/jquery.scianimator.min.js"></script>
<style type="text/css">
.reveal code.r {
   font-size: 1em
}
img {
  width: 450px;
  height: 300px;
}
</style>

```{r setup, echo = FALSE}

# global.par doesn't work
set.par <- function(...) {
  par(mar = c(3.1, 3.1, 1.1, 1.1), xaxt = 'n', yaxt = 'n')
  return(invisible(NULL))
}
knitr::knit_hooks$set(set.par = set.par)
# nb setting fig.show = 'animate' here doesn't seem to work
knitr::opts_chunk$set(interval = 0.25, out.height = '25%', fig.width = 4, fig.height = 2.5)
knitr::opts_knit$set(animation.fun = knitr::hook_scianimator, global.par = TRUE)

```

## Simple example

- NB: I set `par(xaxt = 'n', yaxt = 'n')` to remove clutter

```{r, fig.show='animate', set.par=TRUE}
library(anim.plots)
anim.plot(1:5, 1:5, col = "green")

```

## `times` controls the frames

```{r, fig.show='animate', set.par=TRUE}
x <- rep(1:100/10, 20) 
times <- rep(1:20, each = 100) # twenty frames with 100 points each
y <- sin(x*times/4)
anim.plot(x, y, times = times, type = "l", col = "orange", lwd = 2)
```

## Incremental animations

- Use `window`:
```{r, fig.show='animate', set.par=TRUE}
symbs <- anim.plot(rep(1:10, 2), rep(2:1, each = 10), window = 1:t, pch = 1:20, ylim = c(0,3), cex = 2,
      col = 1:5, xlab = paste("Plot symbols", 1:20))
```


## Parameters

- Per-point parameters get recycled to the number of points
- Per-plot parameters get recycled to the number of plot frames
- For more complex parameters, use a matrix

## Example - zooming in

```{r, fig.show='animate', set.par=TRUE}
x <- rnorm(4000); y<- rnorm(4000)
x <- rep(x, 40); y <- rep(y, 40)
xlims <- 4*2^(-(1:40/10))
ylims <- xlims <- rbind(xlims, -xlims) 
anim.plot(x, y, times=40, xlim=xlims, ylim=ylims, col=rgb(0,1,.5,.3), pch=19, 
      xaxt = 'n', yaxt = 'n')
```


## Moving window

- Use `window` again:
```{r, fig.show='animate', set.par=TRUE}
## discoveries 1860-1959
xlim <- rbind(1860:1959,1870:1969)
anim.plot(1860:1959, discoveries, times = 1:100, xlim = xlim, col = "red", 
      xaxp = rbind(xlim, 10), window = t:(t+10), type = "h", lwd = 8, 
      xaxt = "s", yaxt = "s")
```

## Formula interface

```{r, fig.show='animate', set.par=TRUE}
data(ChickWeight)
ChickWeight$chn <- as.numeric(as.factor(ChickWeight$Chick))
tmp <- anim.plot(weight ~ chn + Time, data = ChickWeight, col = as.numeric(Diet), 
      pch = as.numeric(Diet))
```


## Run extra commands

- Use `replay` with `before` and `after`:
```{r, fig.show='animate', set.par=TRUE}
replay(tmp, after = legend("topleft", legend = paste("Diet", 1:4), pch = 1:4, col = 1:4, horiz = TRUE))
```

# Plots available

## Histogram

```{r, fig.show='animate', set.par=TRUE}
anim.hist(rep(rnorm(5000), 7), times = rep(1:7, each = 5000), 
     breaks = c(5, 10, 20, 50, 100, 200, 500, 1000), main = "Normal distribution")
```

## Animated curve

```{r, fig.show='animate', set.par=TRUE}
anim.curve(x^t, times = 10:50 / 10, n = 20)
```

## Contour plot

- Using a dataset I made earlier

```{r, echo=FALSE}
data(volcano)
tmp <- volcano
tmp[] <- 200 - ((row(tmp) - 43)^2 + (col(tmp) - 30)^2) / 20
cplot <- array(NA, dim = c(87, 61, 20))
cplot[ , , 1] <- tmp
cplot[ , , 20] <- volcano
cplot <- apply(cplot, 1:2, function(x) seq(x[1], x[20], length.out = 20))
cplot <- aperm(cplot, c(2, 3, 1))
```

```{r, fig.show='animate', set.par=TRUE}
anim.contour(z = cplot, times = 1:20, levels = 80 + 1:12 * 10, lty = c(1, 2, 2))
```

## 3D animation

```{r, echo=FALSE}
cplot2 <- apply(cplot, 1:2, function(x) seq(0, x[20], length.out = 20))
cplot2 <- aperm(cplot2, c(2, 3, 1))
```

```{r, fig.show='animate', set.par=TRUE}
par(mar = rep(1.1, 4))
anim.persp(z = cplot2, times = 1:20, xlab = "", ylab = "", zlab = "Height", phi = 45,
      theta = 30, speed = 5, border = NA, r = 3, col = "yellowgreen", 
      shade = .5, box = FALSE)
```


# Thanks!

- http://github.com/hughjonesd/anim.plots


